% ==========================================================================
%  Thesis Section -- The statesnap R Package
%
%  Standalone section describing the reusable R package prototype that
%  generalises ShinySwarm's checkpoint mechanism (the library is named statesnap). This is thesis-ready prose
%  (Discussion / Future Work material) describing WHAT the artifact is and
%  WHY it matters. It is deliberately separate from thesis-notes.tex, which
%  records HOW the library was built and is kept for reproducibility.
% ==========================================================================
\documentclass[11pt,a4paper]{article}
\usepackage[margin=2.5cm]{geometry}
\usepackage{booktabs,tabularx}
\usepackage{hyperref}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{listings}
\usepackage{xcolor}
\lstset{
  basicstyle=\ttfamily\small,
  breaklines=true,
  columns=fullflexible,
  keepspaces=true,
  frame=single,
  framesep=4pt,
  xleftmargin=4pt
}

\title{The \texttt{statesnap} R Package:\\
  A Reusable, Transport-Agnostic Full-State Capture and Restore Library}
\author{Iva Gunjaca\\Master of Software Engineering, University of Amsterdam}
\date{\today}

\begin{document}
\maketitle

% --------------------------------------------------------------------------
\section{Overview}
% --------------------------------------------------------------------------

To validate that ShinySwarm's checkpoint mechanism generalises beyond the
specific microservice architecture in which it was developed, a standalone R
package, \texttt{statesnap}, was implemented as a prototype. The package
distils the full-state capture-and-restore pattern --- previously implemented
individually inside each of the eight Shiny applications --- into two functions
that any Shiny application can use, with no dependency on Spring Boot, Redis,
Kafka, or Kubernetes.

The package is named \texttt{statesnap} to distinguish the reusable library from
\emph{ShinySwarm}, the collaborative microservice system that is the subject of
this thesis. Although it was extracted from that system and is demonstrated with
Shiny, the library does not depend on Shiny: its input and accessor objects are
duck-typed, so the core works with any list-like collection of values and any
getter/setter function. Shiny is therefore a suggested, not a required,
dependency.

The package is a proof of concept. Its purpose in this thesis is to demonstrate
the \emph{feasibility and generality} of the design proposed in the future-work
notes: that exact, reproducible state sharing for interactive applications
can be expressed as a small, transport-agnostic library. The prototype passes
\texttt{R~CMD~check} without errors, warnings, or notes, and is covered by an
automated test suite (Section~\ref{sec:testing}).

% --------------------------------------------------------------------------
\section{Motivation}
% --------------------------------------------------------------------------

Every application in ShinySwarm follows the same checkpoint pattern: capture the
session state (inputs plus computed reactive values), serialise it to JSON, and
later restore it by pushing the values back into the running application. In the
eight-application suite this logic was duplicated, with each application manually
building JSON payloads using application-specific field names.

Existing Shiny state-sharing mechanisms --- \texttt{shinyURL} and the built-in
bookmarking facility --- transfer only \emph{inputs}, forcing the receiving
session to recompute every output. For deterministic analyses this merely costs
time. For non-deterministic analyses, such as a Monte Carlo simulation executed
without \texttt{set.seed()}, recomputation produces \emph{different} results, so
input-only sharing cannot reproduce what the original user saw.

The \texttt{statesnap} package addresses both problems. It standardises the
duplicated logic into two function calls, and it captures computed reactive
values \emph{alongside} inputs, so the restored session reproduces the original
outputs exactly rather than recomputing them.

% --------------------------------------------------------------------------
\section{Design}
% --------------------------------------------------------------------------

\subsection{Transport agnosticism}

The central design decision is that the library produces and consumes JSON
strings only. It contains no networking, persistence, or broker logic. How a
checkpoint is transmitted or stored --- a REST request, a Kafka message, a Redis
entry, a local file, a database row --- is left entirely to the caller. The same
\texttt{capture\_state()} output is therefore usable in every ShinySwarm
architecture and in a plain single-user Shiny application running on a laptop.

\subsection{Public interface}

The package exposes two core functions and two typed wrappers:

\begin{lstlisting}[language=R]
capture_state(input, ..., max_size = NULL, compress = TRUE)
restore_state(session, state_json, ..., file_dir = tempdir(),
              allow_rds = FALSE, max_size = NULL)
state_file(path)   # wrap a file to embed
state_rds(obj)     # wrap an R object to serialise
\end{lstlisting}

Because R's reactive system exposes no mechanism to discover
\texttt{reactiveVal} or \texttt{reactiveValues} objects automatically, the
reactive values to capture or restore are registered explicitly by name through
\texttt{...}. This is a deliberate consequence of Shiny's design, not a
limitation of the library.

\subsection{Typed wrappers}

A Shiny application holds several kinds of state, each requiring a different
capture strategy. Rather than proliferating functions, the library uses
type-hinted wrappers that tell \texttt{capture\_state()} how to handle each
piece of state. The result is always a single JSON string; the wrappers control
what goes inside it.

\begin{center}
\begin{tabular}{lll}
\toprule
\textbf{State type} & \textbf{Example} & \textbf{Capture strategy} \\
\midrule
Scalar input        & \texttt{input\$num1 = 42}    & stored as-is \\
Reactive value      & \texttt{result()}             & accessor called, value stored \\
Reactive dataframe  & \texttt{shared\_df()}         & converted to a JSON array \\
File reference       & \texttt{summary.csv}          & \texttt{state\_file()}: read, gzip, base64 \\
Binary R object      & trained model                 & \texttt{state\_rds()}: serialise, gzip, base64 \\
\bottomrule
\end{tabular}
\end{center}

This generalises the Climate Anomaly checkpoint enrichment, which in the
microservice implementation embeds a processed CSV file into the checkpoint. In
the architecture that logic lives in the Spring Boot backend; in the library it
reduces to a single \texttt{state\_file()} call inside the application's save
handler, removing the need for the backend to mount shared volumes or parse CSV.

\subsection{Usage example}

With the library, an application's save and restore logic each reduce to a single
call:

\begin{lstlisting}[language=R]
library(shiny)
library(statesnap)

server <- function(input, output, session) {
  result <- reactiveVal(0)
  observeEvent(input$calculate, result(input$num1 + input$num2))
  output$result <- renderText(result())

  observeEvent(input$save_btn, {
    json <- capture_state(input, result = result)
    writeLines(json, "checkpoint.json")   # transport chosen by caller
  })

  observeEvent(input$restore_btn, {
    json <- paste(readLines("checkpoint.json"), collapse = "\n")
    restore_state(session, json, result = result)
  })
}
\end{lstlisting}

Inputs are restored through Shiny's low-level \texttt{session\$sendInputMessage()}
mechanism; reactive values are set directly through their accessors.

% --------------------------------------------------------------------------
\section{Reproducibility Guarantee}
% --------------------------------------------------------------------------

The package's defining property is exact reproduction of non-deterministic
results. Because computed reactive values are stored in the checkpoint rather
than recomputed on restore, a value produced by a stochastic process is
preserved bit-for-bit. To make this hold, the JSON serialisation is configured
to retain full numeric precision (\texttt{jsonlite::toJSON(..., digits = NA)});
the default four-decimal rounding would silently break the guarantee. This was
confirmed by a test in which a Monte Carlo mean computed without a fixed seed is
captured, the live value is then changed to a different draw, and a restore
recovers the original value exactly.

% --------------------------------------------------------------------------
\section{Security Considerations}
% --------------------------------------------------------------------------

Building the prototype surfaced a consideration that is absent from input-only
sharing approaches and that connects directly to the cross-user restore scenario
of RQ5: \textbf{a checkpoint is untrusted input}. When user B restores a
checkpoint authored by user A, the checkpoint must be treated as data from an
untrusted source. The restore path therefore defends against three classes of
attack.

\begin{table}[h]
\centering
\begin{tabularx}{\textwidth}{lX}
\toprule
\textbf{Attack} & \textbf{Defence} \\
\midrule
Path traversal via a crafted embedded filename
  & Filenames are reduced to a safe basename; path separators, drive letters,
    home-directory expansion, and \texttt{..} are rejected, so a restored file
    cannot escape its target directory. \\[4pt]
Arbitrary code execution via \texttt{unserialize()}
  & Restoring a serialised object can execute code while the object is
    reconstructed. \texttt{state\_rds} restoration is therefore disabled by
    default (\texttt{allow\_rds = FALSE}) and must be explicitly enabled for
    trusted checkpoints. \\[4pt]
Memory exhaustion (``decompression bomb'')
  & A small compressed payload can expand to many gigabytes. Decompression is
    performed as a bounded stream that aborts as soon as the running output size
    exceeds \texttt{max\_size}, so peak memory stays bounded regardless of the
    claimed decompressed size. \\
\bottomrule
\end{tabularx}
\caption{Security defences in \texttt{restore\_state()}.}
\end{table}

This is a research observation in its own right: full-state sharing trades the
recomputation cost of input-only sharing for a serialisation surface that must
be treated as a trust boundary. Any system that shares computed state across
users inherits this obligation.

% --------------------------------------------------------------------------
\section{Testing and Verification}
\label{sec:testing}
% --------------------------------------------------------------------------

The package is verified at two levels. Unit tests exercise the capture/restore
logic directly, including serialisation round-trips, the typed wrappers, the
exact-reproducibility property for non-deterministic values, and the three
security defences (path traversal, the \texttt{allow\_rds} gate, size limits, and
the decompression-bomb guard). An integration test runs the save/restore cycle
against Shiny's actual reactive engine using \texttt{shiny::testServer()},
confirming the library works with real \texttt{reactiveVal} and
\texttt{reactiveValues} objects rather than test doubles.

The package passes \texttt{R~CMD~check} with a clean result (no errors, warnings,
or notes), the standard correctness bar for an R package.

% --------------------------------------------------------------------------
\section{Performance}
\label{sec:performance}
% --------------------------------------------------------------------------

The library's cost is the in-process work of serialising and deserialising
state. This is conceptually distinct from the system-level benchmarks reported
elsewhere in this thesis, which measure HTTP, database, and session round-trips
under load: the library's cost is the serialisation; the transport's cost is
everything around it. The two should not be conflated. The benchmarks below were
produced by \texttt{inst/benchmarks/benchmark.R}, which uses the
\texttt{microbenchmark} package and runs headless via \texttt{Rscript}.

\paragraph{Methodology and caveats.} Each case is timed over many iterations
(the script defaults to 100; the illustrative run reported here used 50) and the
\emph{median} is reported with the interquartile range (IQR), so that
garbage-collector jitter does not distort the result. All figures
were collected on a single machine with a warm cache and therefore characterise
relative behaviour and scaling rather than absolute deployment latency. The
numbers in Tables~\ref{tab:bench-latency}--\ref{tab:bench-tradeoff} are
illustrative of a representative run; the script regenerates them, together with
two plots, on any machine.

\subsection{Latency versus payload size}

Capture and restore latency scale approximately linearly with the size of the
state, dominated by JSON serialisation and parsing. Small checkpoints (the
common case for interactive applications, on the order of a few kilobytes)
complete in well under a millisecond; multi-megabyte checkpoints cost tens to
hundreds of milliseconds.

\begin{table}[h]
\centering
\begin{tabular}{lrrr}
\toprule
\textbf{State size} & \textbf{JSON bytes} & \textbf{Capture (ms)} & \textbf{Restore (ms)} \\
\midrule
1 KB    & 1{,}398      & 0.66  & 0.07  \\
10 KB   & 13{,}279     & 0.87  & 0.35  \\
100 KB  & 133{,}694    & 3.53  & 3.66  \\
1 MB    & 1{,}361{,}546  & 30.52 & 33.10 \\
5 MB    & 6{,}878{,}022  & 292.71 & 182.13 \\
10 MB   & 13{,}866{,}628 & 707.34 & 385.50 \\
\bottomrule
\end{tabular}
\caption{Median capture/restore latency by payload size (illustrative run,
50 iterations/case). Values scale roughly linearly with state size.}
\label{tab:bench-latency}
\end{table}

\subsection{Compression effectiveness}

Payloads are gzip-compressed before base64 encoding, which offsets base64's
roughly one-third size inflation. The benefit depends heavily on the
redundancy of the data: highly repetitive text (such as a tabular CSV) compresses
by orders of magnitude, structured data frames compress moderately, and
already-dense binary objects (such as a serialised model) compress least.

\begin{table}[h]
\centering
\begin{tabular}{lrrrr}
\toprule
\textbf{Case} & \textbf{Raw bytes} & \textbf{Gzip bytes} & \textbf{Ratio} & \textbf{Saved} \\
\midrule
CSV file (repetitive)   & 1{,}266{,}834 & 3{,}297     & 384.24$\times$ & 99.7\% \\
data.frame (\texttt{state\_rds}) & 1{,}383{,}879 & 549{,}150 & 2.52$\times$ & 60.3\% \\
\texttt{lm} model (\texttt{state\_rds}) & 4{,}003{,}695 & 2{,}555{,}874 & 1.57$\times$ & 36.2\% \\
\bottomrule
\end{tabular}
\caption{Compression ratio by data type (illustrative run). Repetitive text
benefits most; dense binary objects least.}
\label{tab:bench-compression}
\end{table}

\subsection{Full-state versus input-only}

The defining trade-off of the design can be quantified directly. Input-only
sharing transfers a handful of bytes but cannot reproduce computed outputs;
full-state sharing transfers the computed result as well, at a proportional cost
in size and time, and in return reproduces non-deterministic results exactly.

\begin{table}[h]
\centering
\begin{tabular}{lrrc}
\toprule
\textbf{Approach} & \textbf{JSON bytes} & \textbf{Capture (ms)} & \textbf{Reproduces non-det.\ output} \\
\midrule
input-only  & 137        & 0.41  & no  \\
full-state  & 1{,}361{,}864 & 28.81 & yes \\
\bottomrule
\end{tabular}
\caption{Full-state versus input-only for a representative application
(inputs plus a 1~MB computed result; illustrative run). The size and time
overhead is the measurable price of exact reproducibility.}
\label{tab:bench-tradeoff}
\end{table}

For this case, full-state capture costs substantially more in both bytes and time
than input-only sharing. This overhead is not a defect but the explicit cost of
the guarantee the library provides: input-only approaches are cheaper precisely
because they discard the computed state that full-state capture exists to
preserve. The relevant comparison is therefore not ``cheaper versus more
expensive'' but ``recomputed (and, for non-deterministic analyses, irreproducible)
versus reproduced exactly''.

% --------------------------------------------------------------------------
\section{Limitations and Further Work}
% --------------------------------------------------------------------------

The prototype has known limitations, several of which are inherent to Shiny's
design rather than to the library:

\begin{itemize}
  \item \textbf{File-input widgets} cannot be restored through
    \texttt{sendInputMessage()}; the \texttt{state\_file()} wrapper embeds the
    uploaded content as a workaround.
  \item \textbf{Shiny modules} use namespaced input identifiers (for example,
    \texttt{panel1-threshold}). Because \texttt{capture\_state()} reads
    \texttt{names(input)} at the top level and \texttt{restore\_state()} pushes
    values back through the parent \texttt{session}, the library does not model
    the namespace hierarchy that modules introduce, and restoring across module
    session boundaries is not yet supported. Supporting it would require either
    namespace-aware capture (recording each value's namespace path) or recursive
    capture and restore that is invoked from within each module's own
    \texttt{session}, and merging the resulting fragments. This was deliberately
    left out of scope for the prototype: the applications studied in this thesis
    are flat (non-modular), so module support would address a case that does not
    arise in the evaluation. It is documented here so that a future user who
    adopts the library in a modular application is aware of the boundary.
  \item \textbf{Explicit registration} of reactive values is required, because R
    provides no API to enumerate them.
  \item \textbf{\texttt{reactiveValues} keys cannot be deleted} in Shiny; on
    restore, keys absent from the checkpoint are cleared to \texttt{NULL} so
    their values do not leak, rather than being removed entirely.
\end{itemize}

Natural extensions include support for Shiny modules through recursive
capture/restore across session hierarchies (Section~\ref{sec:performance}
benchmarks the flat case only), extending the micro-benchmarks to the specific
state profiles of each studied application, and publication on CRAN to make the
reproducibility contribution available to the wider R~Shiny community without any
microservice infrastructure.

% --------------------------------------------------------------------------
\section{Contribution}
% --------------------------------------------------------------------------

The \texttt{statesnap} package shows that the reproducibility mechanism
developed for the ShinySwarm architecture is not tied to that architecture. The
same guarantee --- exact reproduction of computed results across users,
machines, and time --- can be delivered by a minimal, transport-agnostic library
usable by any Shiny developer. In doing so it both validates the design proposed
as future work and isolates a general security obligation for any system that
shares computed application state.

\end{document}
