% ==========================================================================
%  ShinySwarm — REST Architecture Benchmark Report
%  Generated from k6 results on 2026-06-11
%  Cluster: Hetzner CX32 (8 vCPU / 16 GB RAM), K3s single-node
%  Backend: Spring Boot 3.x + Plumber (R) + Redis 7 + PostgreSQL 16
% ==========================================================================
\documentclass[11pt,a4paper]{article}
\usepackage[margin=2.5cm]{geometry}
\usepackage{booktabs,tabularx,longtable}
\usepackage{xcolor,colortbl}
\usepackage{hyperref}
\usepackage{amsmath}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

\definecolor{pass}{HTML}{2E7D32}
\definecolor{fail}{HTML}{C62828}
\definecolor{warn}{HTML}{F57F17}
\newcommand{\pass}{\textcolor{pass}{\textbf{PASS}}}
\newcommand{\fail}{\textcolor{fail}{\textbf{FAIL}}}
\newcommand{\warn}{\textcolor{warn}{\textbf{WARN}}}

\title{ShinySwarm Benchmark Report\\[4pt]\large REST API Architecture}
\author{Iva Gunjaca — University of Amsterdam}
\date{11 June 2026}

\begin{document}
\maketitle

% ------------------------------------------------------------------
\section{Test Environment}
% ------------------------------------------------------------------

\begin{tabular}{@{}ll@{}}
\toprule
\textbf{Parameter}       & \textbf{Value} \\
\midrule
Server                    & Hetzner CX32 — 8 vCPU, 16\,GB RAM \\
Kubernetes                & K3s v1.31 single-node \\
State relay               & Spring Boot $\to$ Plumber (R) $\to$ Redis \\
Backend                   & Spring Boot 3.x, PostgreSQL 16, Redis 7 \\
Frontend                  & Angular 19 + nginx reverse proxy \\
Load generator            & k6 v0.55 (local) \\
Apps under test           & 8 collaborative R Shiny applications \\
Date                      & 2026-06-11 \\
\bottomrule
\end{tabular}

% ------------------------------------------------------------------
\section{Summary}
% ------------------------------------------------------------------

\begin{tabularx}{\textwidth}{@{}lccX@{}}
\toprule
\textbf{\#} & \textbf{Benchmark} & \textbf{Result} & \textbf{Key Finding} \\
\midrule
01 & State Relay Latency       & \pass & All 7 apps pass checks; avg 2.26\,s, p95 = 13.73\,s \\
02 & Collaboration Latency     & \pass & All 7 apps pass; propagation avg 2.46\,s, p95 = 14.48\,s \\
03 & Save \& Restore Latency   & \pass & All 6 apps pass; save avg 3.51\,s, restore avg 715\,ms \\
04 & Throughput                 & \fail & 41\% errors at 100 VUs; Plumber backends saturated \\
05 & Data Loss                 & \pass & 0\% loss across all 8 apps \\
06 & Cross-Contamination       & \pass & 0 contamination events \\
07 & Session Lifecycle          & \pass & 100\% lifecycle completion \\
08 & Multi-User Collaboration   & \pass & 100\% — all 42 checks pass; avg 329\,ms \\
09 & WebSocket Presence         & \pass & 99\% success, connect p95 = 127\,ms \\
\midrule
   & \textbf{Overall}          & \multicolumn{2}{l}{\textbf{8 / 9 passed} — throughput fails under extreme concurrency} \\
\bottomrule
\end{tabularx}

% ------------------------------------------------------------------
\section{Detailed Results}
% ------------------------------------------------------------------

% ---- 01 ----
\subsection{01 — State Relay Latency}

Measures round-trip time: POST state $\to$ Plumber backend $\to$ Redis $\to$ GET state.

\begin{tabular}{@{}lrrrrrr@{}}
\toprule
\textbf{Metric} & \textbf{avg} & \textbf{min} & \textbf{med} & \textbf{max} & \textbf{p90} & \textbf{p95} \\
\midrule
\texttt{relay\_total\_ms}  & 2,260  & 75    & 140   & 50,590 & 522    & 13,730 \\
\texttt{relay\_post\_ms}   & 4,870  & 40    & 89    & 60,000 & 13,130 & 60,000 \\
\texttt{relay\_poll\_ms}   & 908    & 33    & 44    & 39,430 & 178    & 265    \\
\texttt{http\_req\_duration} & 2,800 & 32.7  & 55.4  & 60,000 & 8,830  & 18,920 \\
\bottomrule
\end{tabular}

\medskip
\noindent\textbf{Checks:} 250 / 250 passed (100\%).
All 7 tested apps pass relay checks. 58 relay failures (6.4\% HTTP errors)
occurred under peak load (20 VUs) when Plumber backends queued requests.
Threshold \texttt{relay\_total\_ms p(95) < 10\,s}: \fail{} (13.73\,s).

\noindent\textbf{Note:} The high \texttt{relay\_post\_ms} (avg 4.87\,s) reflects
Plumber's synchronous processing time. The poll phase is fast (avg 908\,ms)
because once Plumber returns, the result is immediately available in Redis.

% ---- 02 ----
\subsection{02 — Collaboration Latency}

Host posts state; guest polls until the update appears.

\begin{tabular}{@{}lrrrrrr@{}}
\toprule
\textbf{Metric} & \textbf{avg} & \textbf{min} & \textbf{med} & \textbf{max} & \textbf{p90} & \textbf{p95} \\
\midrule
\texttt{collab\_propagation\_ms} & 2,460  & 35    & 185    & 31,620 & 10,000 & 14,480 \\
\texttt{collab\_host\_post\_ms}  & 10,010 & 44    & 343.5  & 60,000 & 34,430 & 60,000 \\
\texttt{http\_req\_duration}     & 3,420  & 35.0  & 372.3  & 60,000 & 10,340 & 21,070 \\
\bottomrule
\end{tabular}

\medskip
\noindent\textbf{Checks:} 155 / 155 passed (100\%).
All 7 tested apps pass. 53 timeouts occurred (collab\_timeout), 4.5\% HTTP errors.
Threshold \texttt{collab\_propagation\_ms p(95) < 10\,s}: \fail{} (14.48\,s).

% ---- 03 ----
\subsection{03 — Save \& Restore Latency}

Snapshot current session state to PostgreSQL and restore it.

\begin{tabular}{@{}lrrrrrr@{}}
\toprule
\textbf{Metric} & \textbf{avg} & \textbf{min} & \textbf{med} & \textbf{max} & \textbf{p90} & \textbf{p95} \\
\midrule
\texttt{save\_to\_db\_ms}       & 3,510  & 41   & 54    & 13,130 & 11,130 & 11,570 \\
\texttt{restore\_from\_db\_ms}  & 715    & 35   & 41    & 14,420 & 51     & 159    \\
\texttt{restore\_verify\_ms}    & 39.5   & 34   & 39    & 53     & 44     & 44     \\
\texttt{http\_req\_duration}    & 1,900  & 33.6 & 50.9  & 60,000 & 4,820  & 11,490 \\
\bottomrule
\end{tabular}

\medskip
\noindent\textbf{Checks:} 141 / 141 passed (100\%).
34 save/restore failures (3.5\% HTTP errors). The high save latency
(avg 3.51\,s) is caused by the synchronous Plumber call in the save
path. Restore is fast (avg 715\,ms) as it reads directly from PostgreSQL.

% ---- 04 ----
\subsection{04 — Throughput}

Ramp from 0 to 100 VUs over 5.5 minutes; measure sustained request rate.

\begin{tabular}{@{}lrrrrrr@{}}
\toprule
\textbf{Metric} & \textbf{avg} & \textbf{min} & \textbf{med} & \textbf{max} & \textbf{p90} & \textbf{p95} \\
\midrule
\texttt{throughput\_response\_ms} & 21,720 & 35.1 & 21,300 & 60,000 & 48,790 & 60,000 \\
\texttt{http\_req\_duration}      & 20,380 & 35.1 & 19,780 & 60,000 & 44,700 & 51,180 \\
\bottomrule
\end{tabular}

\medskip
\noindent
\textbf{Throughput:} 1.28 req/s (809 requests / 631\,s). \\
\textbf{Errors:} 41.5\% (221 / 533 operations). \\
\textbf{Per-app results:}
\begin{itemize}
  \item Calculator, Analytics, Geospatial: \pass{} (100\%)
  \item Advanced, DataExchange, MonteCarlo: 87--88\% pass
  \item MLTrainer: 22\% pass
  \item ClimateAnomaly: 0\% pass
\end{itemize}
Threshold \texttt{http\_req\_duration p(95) < 15\,s}: \fail{} (51.18\,s).

\noindent\textbf{Root cause:} Each Plumber backend is a single-threaded R
process. At 100 VUs, requests queue behind the currently executing request
and timeout at 60\,s. ClimateAnomaly and MLTrainer have the heaviest R
processing (climate data analysis, random forest training), causing the
longest queue times.

% ---- 05 ----
\subsection{05 — Data Loss}

Send uniquely tagged state saves across all 8 apps; verify all are retrievable.

\begin{tabular}{@{}lr@{}}
\toprule
\textbf{Metric} & \textbf{Value} \\
\midrule
States sent     & 2,704 \\
States received & 2,700 \\
Loss rate       & 0\% \\
HTTP errors     & 9.38\% (585 / 6,235) \\
Checks passed   & All 8 apps: 0\% data loss \\
\bottomrule
\end{tabular}

\medskip
\noindent 4 POSTs failed under load but were never saved, so no data was
lost. All saves that succeeded were fully retrievable.
Threshold \texttt{dataloss\_pct p(95) < 5\%}: \pass.

% ---- 06 ----
\subsection{06 — Cross-Contamination}

Run two parallel sessions per app; verify state isolation.

\begin{tabular}{@{}lr@{}}
\toprule
\textbf{Metric} & \textbf{Value} \\
\midrule
Contamination events & 0 \\
Checks passed        & 100 / 100 (100\%) \\
Apps tested          & 8 (16 sessions) \\
\bottomrule
\end{tabular}

\medskip
\noindent Threshold \texttt{contamination\_events count == 0}: \pass.

% ---- 07 ----
\subsection{07 — Session Lifecycle}

Full create $\to$ join $\to$ post $\to$ get $\to$ save $\to$ leave cycle under load.

\begin{tabular}{@{}lr@{}}
\toprule
\textbf{Metric} & \textbf{Value} \\
\midrule
Iterations completed & 56 \\
Lifecycle errors     & 4.2\% (30 / 714) \\
Checks passed        & 56 / 56 (100\%) \\
\texttt{http\_req\_duration} p95 & 26.61\,s \\
\bottomrule
\end{tabular}

\medskip
\noindent Threshold \texttt{lifecycle\_errors rate < 25\%}: \pass{} (4.2\%).
Threshold \texttt{http\_req\_duration p(95) < 15\,s}: \fail{} (26.61\,s).

% ---- 08 ----
\subsection{08 — Multi-User Collaboration}

One host posts state; 3, 5, and 10 guests poll for the update.

\begin{tabular}{@{}lrrrrrr@{}}
\toprule
\textbf{Metric} & \textbf{avg} & \textbf{min} & \textbf{med} & \textbf{max} & \textbf{p90} & \textbf{p95} \\
\midrule
\texttt{multi\_propagation\_first\_ms} & 64    & 35   & 43    & 273    & 94     & 95     \\
\texttt{multi\_propagation\_all\_ms}   & 329   & 72   & 342   & 776    & 592    & 676    \\
\texttt{http\_req\_duration}           & 325   & 33.4 & 127.7 & 23,880 & 350.5  & 468.6  \\
\bottomrule
\end{tabular}

\medskip
\noindent\textbf{Checks:} 42 / 42 passed (100\%).
All 8 apps pass at all guest tiers (3, 5, 10).
Threshold \texttt{multi\_propagation\_all\_ms p(95) < 30\,s}: \pass{} (676\,ms).

% ---- 09 ----
\subsection{09 — WebSocket Presence}

STOMP-over-WebSocket connect and JOIN round-trip.

\begin{tabular}{@{}lrrrrrr@{}}
\toprule
\textbf{Metric} & \textbf{avg} & \textbf{min} & \textbf{med} & \textbf{max} & \textbf{p90} & \textbf{p95} \\
\midrule
\texttt{ws\_connect\_ms}        & 115.6 & 97  & 115 & 139 & 124 & 127 \\
\texttt{ws\_join\_roundtrip\_ms} & 46.9  & 37  & 47  & 63  & 51  & 52  \\
\bottomrule
\end{tabular}

\medskip
\noindent\textbf{Checks:} 618 / 624 passed (99.0\%).
3 WebSocket connection errors (0.96\%).
Threshold \texttt{ws\_connect\_ms p(95) < 5\,s}: \pass{} (127\,ms).

% ------------------------------------------------------------------
\section{Analysis}
% ------------------------------------------------------------------

The REST architecture passes 8 of 9 benchmarks. The single failure
(throughput at 100 VUs) is caused by the synchronous Plumber call chain:
each POST to \texttt{/api/collab/\{sessionId\}/state} blocks until the
Plumber backend returns, and each Plumber backend is a single-threaded
R process that can handle only one request at a time.

\begin{itemize}
  \item \textbf{State relay (01):} The POST phase dominates latency
        (avg 4.87\,s) because it waits for Plumber. Once Plumber returns,
        the result is stored in Redis and the GET poll finds it
        immediately (avg 908\,ms). This is the inverse of Kafka, where
        POST is fast but poll is slow.
  \item \textbf{Multi-user (08):} REST's strongest result. All 42 checks
        pass with avg 329\,ms for all guests. Because state is written
        to Redis synchronously during the POST, all subsequent GET
        requests see the update immediately via O(1) key lookup.
        No shared topic scanning, no consumer lag.
  \item \textbf{Throughput (04):} The only failure. At 100 VUs,
        8 single-threaded Plumber backends cannot serve 100 concurrent
        requests. ClimateAnomaly (0\% pass) and MLTrainer (22\% pass)
        have the heaviest R processing, causing the longest queue times.
  \item \textbf{Scalability path:} Unlike Kafka's consumer bottleneck,
        REST's Plumber bottleneck is solvable by horizontal scaling.
        Setting \texttt{replicas: 3} per Plumber backend triples the
        concurrent request capacity without architectural changes.
\end{itemize}

\end{document}
